Developmental trajectories of word recognition: a priming study in bilinguals

### 2023-10-21 manuscript/manuscript.qmd
Update manuscript
Add references and remove HTML manuscript

The formation of a mental lexicon is a critical developmental achievement for infants_ learning words allows infants to retrieve socially relevant concepts from fairly arbitrary linguistic forms embedded in speech.

Learning words allows infants to access the rich world made of concepts through the recognition of fairly arbitrary linguistic forms, i.e., words. This remarkable developmental achievement is

There is evidence of cascaded activation during lexical access.

@chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category. For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the **bee**!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development.

Another important source of evidence in favour of a cascaded activation account comes from studies on implicit naming. @mani2010infant showed that 18 month-old infants implicitly generate the labels of name-known pictures presented in silence. In each trial, infants were presented with prime pictures presented in silence, and then with a target and a distractor picture, both presented side-by-side. Finally, the target picture's label was presented auditorily. The authors manipulated the phonological overlap between the prime label and the target label, so that in half of the trials both labels were phonologically related, sharing phonological onset (*cat*-*cup*), or phonologically unrelated (*ball*-*comb*). Prime and target-distractor pairs were semantically unrelated. Interestingly, infants showed a stronger looking preference for the target picture after phonologically related prime pictures, compared to after phonologically unrelated primes. This suggests that infants implicitly named prime pictures despite such pictures having been presented in silence, and that the phonology of the resulting label interacted with the subsequent auditory recognition of the target label. In particular, the authors found a facilitation effect: target looking preference was stronger in phonologically related primes, indicating that the activated phonological segments of the generated prime label facilitated the subsequent activation of the phonological representation of the target label during recognition.

@mani2011phonological used the same paradigm in 24-months-old toddlers, and also found evidence of implicit naming, but in the opposite direction. After phonologically related primes, infants showed weaker target looking preference than after phonologically unrelated primes. These results interpreted these outcome as the result of the implicitly generated prime label interfering with the subsequent recognition of the (phonologically related) target word. The authors suggested that the shift from facilitation to interference from 18 to 24 months they found in both studies might be the result of a developmental shift in which 18-month-olds' lexicon is not yet organised based on phonological similarity, while 24-month-olds' is. This would lead to the former showing a pre-lexical facilitation effect in which such facilitation occurs via the activation of phonological segments, while the latter would show an interference effect due to lateral links between lexical representation having been established. Both studies by Mani and Plunkett provide evidence of implicit naming of name-known pictures by 18-month old toddlers and older, and that the resulting labels interact with the dynamics of recognition of subsequently presented word-forms.

Evidence from cascaded activation in lexical access is also provided by bilinguals. The bilingual lexicon is language-non selective. There is strong evidence in support of bilinguals activating lexical representations from both languages during both speech production and comprehension. Previous studies on bilingual word production support a cascade account for lexical activation in which the phonology of non-selected lexical representations engages in excitatory or inhibitory dynamics in the lexical selection process.

## Co-activation in adults

How does language co-activation shape lexical development?
Marian and Spivey (1999)
Costa et al. (2000)

## Evidence from co-activation though translation
The role of cognates in lexical processing.

Another source of evidence are studies studies involving phonological priming of non-cognates through translation. [SEE REFERENCES FROM VON HOLZEN 2012]

Using an adaptation of @mani2010infant's implicit naming paradigm, @von2014bilinguals provided evidence that bilingual adults generate implicit labels in both labels for visually fixated pictures. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' N400 ERP components were recorded from target word onset. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition.

@@

Floccia et al. (2020) tested toddlers in a cross-language priming paradigm. A prime word was embedded at the end of a carrier sentence that participants listened to. Then participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. Some critiques:

- Since participants heard the auditory label in one of the languages, priming through translation might be compromised.
- Translation equivalents (Exp. 1) are more strongly related at the semantic level with each other than other pairs of semantically related words that are not translations of each other (table-mesa are more strongly associated than table-silla).
- We used longitudinal participants, so we can examine the developmental trajectories of cross-language lateral inhibitory or excitatory connections better.
- Larger sample size.

Onset vs. offset overlap:

- Wu and Thierry (2011): rhyme

@@

>>>>

::: {.callout-note}
How do  bilinguals keep up with monolinguals?
:::

Bilingual infants receive a relative impoverished linguistic input in each of their languages compared to their monolingual peers. There is not reason to think bilinguals receive an increased amount of total linguistic input, so the fact that such input is divided between at least two languages means that they receive less input from each of their languages [@costa2014does]. The input bilinguals receive is also more complex than that of monolinguals. This is reflected in multiple dimensions. For instance, bilinguals are presented with two different, but possible overlapping phoneme inventories, they may be presented with two different grammatical systems, and often presented with more than one label per referent. This does not keep bilinguals from tuning their perceptual abilities to both phoneme inventories of their native languages, distinguishing grammatical irregularities in both languages, or learning more than one label per object. Bilinguals' developmental trajectories might not always be identical to monolinguals', but they achieve their language acquisition milestones at equivalent ages.

::: {.callout-note}
Developmental trajectories of lexical acquisition.
:::

For instance, previous work on lexical acquisition has provided some insight into how bilinguals' vocabulary size is equivalent to monolingual's across ages, and the possible mechanisms behind this. @hoff2012 found that English-Spanish bilinguals' vocabulary in English was influenced by the words they had already acquired in Spanish. This is consistent with previous accounnts of lexical acquisition in which infants acquire new words through preferential acquisition [@hill2013, @fourtassi2020]. A monograph by @floccia2018introduction provided evidence of a language distance effect, where bilinguals learning two languiages sharing higher lexical similarity knew more words in their less dominant language than those learning two language with lower lexical similarity. Later work by @mitchell2022cognate and @garcia-castro@cognate provided evidence for a more direct facilitation effect of cognateness. Cognates are form-simlar translation equivalents. Cognateness is suggested to facilitate word acquisition through parallel activation: the co-activation of translation equivalents in speech might lead to an faster accumulation of learning instances with words, which ultimately results in an earlier age of acquisition.

::: {.callout-note}
Lexical structure, as reflected by word processing.
:::

It is well established that, across languages, monolingual infants start recognising their first word-forms before their first year of life [@jusczyk1999; @halle1994; @vihman2004], with the earliest age at which there is evidencer of familiar word recognition being 6 months [@bergelson2012]. Bilinguals start recognising words at similar ages. Using a Head-turn Prefernce Procedure and ERP registration,  @vihman2013 found English-Welsh bilinguals to be sensitive to the prevalence of auditorily presented words. Infants' looking times and ERP differed when presented with words frequently marked as "familiar" by caregivers of demogrtaphically equivalent peers (e.g., *nappy*, *thank you*, *bottle*, *apple*, and *naughty*), compared to when presented with rare words (e.g., *Eiffel*, *courtly*, *nettle*, *wacky*, and *juncture*).

::: {.callout-note}
Language-non selective lexical access in bilinguals
:::

A critical property of bilingual language processing, both in adults and infants, is that of non-selective access to the lexicon. When listening to speech, bilinguals activate all lexical candidates from either language consistent with the speech signal. Even more surprising, when speaking, bilinguals activate all potential words consistent with the concepts they want to express.
This effect has been systematically reported in adults [@marian1999activation; @costa2000cognate; @thierry2007brain; @dunabeitia2009masked].

Also in toddlers [@von2012language; @von2014bilinguals] and school-aged children [@persici2019lexical; @bosma2020cognate]. Relatedly, some recent studies have investigated the recognition of mispronounced words comparing monolingual and bilingual toddlers. Overall, these studies do not find differences between monolinguals and bilinguals [@von2019impact; @wewalaarachchi2017vowels]. Evidence for differential processing of cognate and non-cognate words is provided by Ramon-Casas and colleagues who observed that SpanishCatalan 24-month-olds were more prone to accept mispronunciations when the stimuli involved cognate words than when they involved non-cognate words [@ramon2009vowel; @ramon2010non].

Although recent work has provided strong eviodence of parallel activation in early ages, how cognateness shapes that organisation of the early bilingual lexicon, and how this is reflected on the dynamics of lexical access during word comprehension and production is unclear.

Using the experimental paradigm developed by [@arias2009lexical; @arias2013s; @singh2014one] compared within and across language
priming effects in Chinese-English toddlers. @kuipers2015bilingualism compared ERP responses to picture-word pairs in English monolinguals and Welsh-English bilinguals where the words were either semantically related or unrelated to a pictured object. Although both studies tested 30-month-olds, the conclusions the authors reached do not fully agree. @singh2014one concluded that there is a discontinuity between adults and young children in the way the two lexicons interact, while @kuipers2010event and @kuipers2015bilingualism concluded that there are no differences in adults’ and toddlers’ lexical-semantic processing.

There is evidence of cascaded activation during lexical access. @chow2017spoken adapted the Visual World Paradigm from @huettig2007tug to explore 24- to 30-month-old toddlers' visual fixation patterns during a word recognition when presented with phonological and semantic distractors. In each trial, the authors presented participants with four semantically and phonologically pictures. Four seconds after pictures onset, a word-form was auditorily presented. The word did not refer to any of the pictures displayed on the screen, but was phonologically related to one of them (both labels shared phonological onset), and semantically related to another one of the pictures (both referents belonged to the same taxonomic category). For instance, participants might be presented with the pictures of a sandwich, a bus, a cat, and a dress. Then they would hear the carrier phrase "Look at the **bee**!". The authors registered participants fixations to the phonological and semantic distractors, and found evidence of a preference for the phonological distractor at earlier stages of the post-naming phase, and a preference for the semantic distractor at later stages of the trial. These results support a cascaded activation account of lexical access during the first stages of lexical development.

Bilinguals' trajectories of word recognition  [@vihman2013]

@floccia2020translation tested 27-month-old simultaneous bilinguals in a cross-language priming paradigm. The two languages participants were learning were English, and an Additional Language (either Cantonese, Dutch, Fench, German, Greek, Italian, Mandarin, Polish, Portuguese, or Spanish). A prime word was embedded at the end of a carrier sentence that participants listened to. Then participants were presented with the auditory label of the target word, and two pictures were shown side-by-side, the target picture, and a distractor picture. In Study 1, the target word was the translation equivalent of the prime word (*cheese*-*fromage* for an English-French bilingual). In other trials, prime and target words were unrelated (e.g., *sock*-*fromage*). There was no phonological overlap at onset between prime and target or distractor labels in either of the two conditions. Participants were presented with two blocks of related and unrelated trials. In one of the blocks, participants were presented with English prime words and Home Language target words (*sock*-*fromage*), and in the other block they were presented with Additional Language prime words and English target words (*fromage*-*sock*). The authors found a cross-linguistic priming effect, in which participants showed a stronger preference for the target picture in the post-naming phase in related trials, as compared to unrelated trials. This effect was found regardless of the language of the prime (English or Home Language), or the dominance of the language of the prime (i.e., whether the prime word belonged to the dominant language). Target preference exceeded chance level in both conditions. In Study 2, the authors tested a similar group of bilinguals in a cross-linguistic semantic priing task, in which participants were primed with words in English, and tested with semantically related words in the Additional Language (or vive versa). This time, prime and target words were not translation equivalents, butr rather words whose referents shared semantic features. Results from Study  2 were parallel to those from Study 1. The authors found a strong semantic priming effect accross languages, regardless of the language of the prime or the dominance of the language of the prime. Overall these results suggest that bilinguals recognosed target words in both conditions, and benefitted from a cross-linguistic facilitation effect from translation equivalents. The symmetry of such effect across both languages suggests that participants lexical access ocurred in parallel for the two members of translation equivalents. As @floccia2020translation note, the stimuli set of Study 2 contains multiple cognates, especially in languiages with a larger degree of phonological overlap across translation pairs, like English and Dutch. The authors run *post hoc* analyses exploring the effect of cognateness on the cross-language priming effects found, which showed no evidence of a cognateness effect on target preference. However, the cross-language effects reported in Study 2 might still reflect the confounding influence of cognateness, as the resulting effect might have been generatd by an interaction by the phonological and semantic overlap between cognate primes and targets.

Another source of evidence supporting the language non-selective account of lexical access is provided by cross-languistic priming priuming thorught translation paradigms. In line with @floccia2020translation, @von2012language found that bilingual toddlers aged 21-to-43-months, and who were learning German and English, showed a stronger target preference in a German word recognition task after being primed with a phonologically related English word, which shared phonological onset with the prime word (e.g., *slide*-*Klide* [German for *dress*]), compared to when prime and target words were phonologically unrelated (e.g., *bottle*-*Ohr* [German for *ear*]). Interestingly, they also found evidence of such priming effect when participants were prime with an English word which was *not* phonologically related to the German target word, but rather its German translation was (e.g., *leg* [with German trasnlation *Bein*]-*Stein* [German for *stone*]). Altogether, these results indicate that bilingual toddlers activate translation equivalents in parallel during word recognition, and that the phonology of both labels may interact with lexical access in the recognition of subsequet words. Importantly, cross-linguitic priming through translation demonstrates that previous findings of a cognateness facilitation in word production and recognition do not result from the mere phonological overlap between auditorily presented and target labels, but also from the phonological overlap between the prime translation (not presented) and the target.

## Implicit naming

The aforementioned studies use priming-though-translation paradigms. In these paradigms, cross-linguistic priming results from activation spreading from the lexical representation of the the (auditorily presented) prime word to the lexical representation of its translation through semantic associations (e.g., translation equivalence). When prime words are auditorily presented, the task design constraints the naming context to one of the languages. The consequence is a potential asymmtry between the activation dynamics of the lexical represetnation of both labels: while the participant has heard one label, the activation of its translation must spread across semantic links. This might lead to a lower strength, and higher lacenty of the activation of the translation.

In the present study, we use an adaptation of an implicit naming paradigm by @mani2010infant, in which participants are primed with a picture, instead of an auditorily presented label. @mani2010infant showed that 18 month-old monolinguals implicitly generate the labels of name-known pictures presented in silence. In each trial, infants were presented with prime pictures presented in silence, and then with a target and a distractor picture, both presented side-by-side. Finally, the target picture's label was presented auditorily. The authors manipulated the phonological overlap between the prime label and the target label, so that in half of the trials both labels were phonologically related, sharing phonological onset (*cat*-*cup*), or phonologically unrelated (*ball*-*comb*). Prime and target-distractor pairs were semantically unrelated. Interestingly, infants showed a stronger looking preference for the target picture after phonologically related prime pictures, compared to after phonologically unrelated primes. This suggests that infants implicitly named prime pictures despite such pictures having been presented in silence, and that the phonology of the resulting label interacted with the subsequent auditory recognition of the target label. In particular, the authors found a facilitation effect: target looking preference was stronger in phonologically related primes, indicating that the activated phonological segments of the generated prime label facilitated the subsequent activation of the phonological representation of the target label during recognition.

@mani2011phonological used the same paradigm in 24-months-old monolinguals, and also found evidence of implicit naming, but in the opposite direction. After phonologically related primes, infants showed weaker target looking preference than after phonologically unrelated primes. These results interpreted these outcome as the result of the implicitly generated prime label interfering with the subsequent recognition of the (phonologically related) target word. The authors suggested that the shift from facilitation to interference from 18 to 24 months they found in both studies might be the result of a developmental shift in which 18-month-olds' lexicon is not yet organised based on phonological similarity, while 24-month-olds' is. This would lead to the former showing a pre-lexical facilitation effect in which such facilitation occurs via the activation of phonological segments, while the latter would show an interference effect due to lateral links between lexical representation having been established. Both studies by Mani and Plunkett provide evidence of implicit naming of name-known pictures by 18-month old toddlers and older, and that the resulting labels interact with the dynamics of recognition of subsequently presented word-forms.

@@

When extended to the bilingual case, this implicit naming task may offer bilingual participants with an uncontrained naming context in whcih the activation of both labels may occur in a truly cascaded fashion for both languages in parallel: the recognition of the prime object depicted activates the (common) semantic representation for both members of the translation equivalent, and this in turn activates the two available phonological word-forms, one in each language. Using an adaptation of @mani2010infant's implicit naming paradigm, @von2014bilinguals provided evidence that bilingual adults generate implicit labels in both labels for visually fixated pictures. The authors presented German-English bilinguals with 120 prime-target pairs. Primes were presented as familiar pictures in silence. After prime picture offset, target words were presented auditorily. Participants' N400 ERP components were recorded from target word onset. The authors manipulated the phonological relationship between the prime and target word-forms within participants' L1 and across L1 and L2. When prime and target were phonologically identical (*Affe*-*Affe*, German for *monkey*), or similar (*Fahne*-*Sahne*, German for *flag* and *ice-cream*, respectively) within L1, participants' showed a reduced N400 amplitude, compared to when prime and target were phonologically unrelated (*Messer*-*Seil*, German for *knife* and *rope*). Critically, a similar effect was found when prime and target were phonologically related through translation (*Rustsche*-*Kleid*, German for *slide* and *dress*). This suggests that participants activated prime labels in both languages in parallel, and that both labels impacted the dynamics of target words recognition.

The role of cognates in lexical processing.

This study also provides a more suitable design for drawing conslusions about developmental change in word recognition trajectories frot wo reasons. First, it extends the age range of participants, compared to previous studies. Second, participants were tested at three selected age points to ensure a more balanced sample size across ages. Third, the partially longitudinal design allows modelling the performance of participants across testing sessions altogether.

. In order to circumvent the problem of limited vocabulary knowledge
in the non-dominant language, we propose to test participants only in their dominant language, therefore
directing our research to the question of how possessing a second language affects processing of the
dominant one in bilingual toddlers [@costa2014does provide a fuller description of the
advantages of such approach].

@@

## Summary

## Context

## Facilitation vs. inhibition

## Onsets vs. offsets

## Problems and generalisatility

## Future steps

## Conclusions
